Regression models for Cylindrical data in Psychology

### 2019-01-29 Manuscript/ManuscriptMarkdownCV.Rmd
Update manuscript

author: 
  - name          : "Jolien Cremers"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "joliencremers@gmail.com"
  - name          : "Helena J.M. Pennings"
    affiliation   : "2,3"
  - name          : "Christophe Ley"
    affiliation   : "4"

@@

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

@@

In social sciences the use of cylindrical data is very common. Such data consist
of a linear and a circular component. @gurtman2011reasoning refers to such data
as vectors, with a directional measure (i.e., the circular component) and a
measure indicating the magnitude (i.e., linear component). Many established
models in psychology are often referred to as circular or circumplex models, but
those models are cylindrical. Examples of such cylindrical models are the
interpersonal circle/circumplex
[@leary1957;@wiggins1996history;@wubbels2006interpersonal], the circumplex of
affect [@russell1980circumplex], the circumplex of human emotion
[@plutchik1997general] or the model of human values [@schwartz1992values].
\newline
\indent Also, many of the more recent types of data that are studied in psychology are
cylindrical. For example, research on human navigation uses data where distance
(i.e., linear component) and direction (i.e., the linear component) are of
interest [@chrastil2017rotational] or in eye-tracking, the saccade data also
consist of both the direction (i.e., the circular variable) and the duration
(i.e., the linear variable)(e.g., @rayner200935th). Apart from the social
sciences data with a circular and linear outcome more commonly occur in
meteorology meteorology [@garcia2013exploring], ecology [@garcia2014test] or
marine research [@lagona2015hidden]
\newline
\indent Up until now researchers studying cylindrical data had to rely on the linear
statistical methods to analyze their research results. Focusing on only the
circular variable or only on the linear variable. However, lately more and more
of these researchers acknowledge that linear methods are not sufficient and call
for new methods [@gurtman2011reasoning; @pennings2017complexity;
@wright2009integrating] that take into account both the circular and the linear
component of these data.
\newline
\indent The aim in the present paper is twofold. Firstly, we intend to fill this gap in
the literature on cylindrical models by adapting four existing cylindrical
models in such a way that they include a regression of both the linear and
circular component of a cylindrical variable onto a set of covariates. From now
on we will therefore refer to the components of the cylindrical variable as
outcome components. Secondly, we will show how using these models can benefit
the analysis of circumplex data and cylindrical data in psychology in general.
More specifically we will show these benefits for interpersonal teacher data
from the field of educational psychology. Apart from modelling the dependence
between the linear and circular component of a cylindrical variable we would
also like to predict the two components from a set of covariates in a regression
model. In our analysis, we will show how this can be done in four different ways
using the adapted cylindrical models.

@@

Thus, data that consist of a variable and a circular variable is called
cylindrical data. A circular variable is different from the linear variable in
the sense that it is measured on a different scale. Figure \ref{circline} shows
the difference between a circular scale (right) and a linear scale (left). The
most important difference is that on a circular scale the datapoints 0\(^\circ\)
and 360\(^\circ\) are connected and in fact represent the same number while on a
linear scale the two ends, \(-\infty\) and \(\infty\), are not connected and
consequently the values 0\(^\circ\) and 360\(^\circ\) are located on different
places on the scale. Both circular data and cylindrical data require special
analysis methods due to this periodicity in the scale of a circular variable
(see e.g. @fisher1995statistical for an introduction to circular data and
@mardia2000directional, @jammalamadaka2001topics and @ley2017modern for a more
elaborate overview).\newline
\indent In the literature, several methods have been put forward to model the
relation between the linear and circular component of a cylindrical variable.
Some of these are based on regressing the linear component onto the circular
component using the following type of relation: \[y = \beta_0 +
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is the
linear component and \(\theta\) the circular component [@mardia1978model;
@johnson1978some; @mastrantonio2015bayesian]. Others model the relation in a
different way, e.g. by specifying a multivariate model for several linear and
circular variables and modelling their covariance matrix
[@mastrantonio2018joint] or by proposing a joint cylindrical distribution. For
example, @abe2017tractable introduce a cylindrical distribution based on a
Weibull distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their respective
shape and concentration parameters. However, none of the methods that have been
proposed thus far include additional covariates onto which both the circular and
linear component are regressed.

@@

(vertical) and Communion (horizontal), see Figure \ref{QTI}. Different parts of
this space are characterized by different teacher behavior, e.g. 'helpful' or
'uncertain'. This two-dimensional space is called the interpersonal
circle/circumplex (IPC). The IPC is ``a continuous order with no beginning or
end'' [@gurtman2009exploring, p. 2]. We call such ordering a circumplex ordering
and the IPC is therefore often called the interpersonal circumplex. The ordering

@@

scores; their respective 0 scores form the origin. Agency and Communion are
measured on a scale from 1 to 5 and for analysis purposes they are later
converted to scale ranging from -1 to 1. Scaling is however only considered an
issue in those instances where cylindrical data is derived from measurements in
bivariate space.}, where \(A\) represents a
score on the Agency dimension and \(C\) represents a score on the Communion
dimension. Note that when placing a unit circle on Figure \ref{QTI} we see that
the Agency dimension is related to the sine of the circular score and the
Communion dimension is related to the cosine of the circular score.

@@

the two dimensions Agency and Communion instead of both dimensions separately
providing us with new insights compared to a separate analysis of the two
dimensions that is standard in the literature (see e.g. @pennings2018interpersonal, @wright2009integrating or @wubbels2006interpersonal). There is however one main drawback: when two-dimensional
data are converted to the circle we lose some information, namely the length of
the two-dimensional vector \((A, C)^t\), \emph{i.e.}, its Euclidean norm
\(\mid\mid (A, C)^t \mid\mid\). This length represents the strength of the
interpersonal behavior a teacher shows towards his/her students. In a
cylindrical model this strength (the linear outcome) can be modeled together
with the type of interpersonal behavior of a teacher (the circular outcome).
This leads to an improved analysis of interpersonal circumplex data, over either
analyzing the two dimensions separately or using a circular model, because we
take all information, circular and linear, into account. In the next section we
introduce several cylindrical models that can be used to analyze the teacher
data. First however we will provide descriptives for the teacher data.

@@

The teacher data was collected between 2010 and 2015 and contains several
repeated measures on the IPC of 161 teachers. Measurements were obtained using
the QTI and taken in different years and classes. For this paper we only
consider one measurement, the first occasion (2010) and largest class if data
for multiple classes were available. This results in a sample of 151 teachers. In addition to the type of interpersonal
behavior (IPC), the circular outcome, and the strength of interpersonal behavior
(IPC strength), the linear outcome, a teachers' self-efficacy (\verb|SE|)
concerning classroom management is used as covariate in the analysis.
In previous research, in psychology and education it had been shown that higher self-efficacy is related to the quality of interpersonal interactions [@locke2007selfefficacy; @want2018selfefficacy]. After listwise
deletion of missings ($3$ in total, only for the self-efficacy) we have a sample
of 148 teachers. Table \ref{Tableteacherdescriptives} shows descriptives for the
dataset. For the circular variable IPC we show sample estimates for the circular
mean $\bar{\theta}$ and concentration $\hat{\rho}$. The circular concentration
lies between 0, meaning the data is not concentrated at all \emph{i.e.} spread
over the entire circle, and 1, meaning all data is concentrated at a single
point on the circle The population values of these parameters are usually, and
also in this paper, referred to as $\mu$ (circular location) and $\kappa$
(circular concentration). For the linear variables (strength IPC and SE) we show
sample estimates of the linear mean and standard deviation. Figure
\ref{dataplot} is a scatterplot showing the relation between the linear and
circular outcome of the teacher data for teachers with low SE (below 1 sd below
the mean), average SE (between 1 sd below and 1 sd above the mean) and high SE
(above 1 sd above the mean).

@@

One of the goals of this paper is to show the benefits of cylindrical methods
for the analysis of circumplex data and cylindrical data in psychology in
general. To do so we decide to focus on four cylindrical models. The models were
selected for their relatively low complexity and the ease with which a
regression structure could be incorporated. But also because they show different
ways of modelling the the linear and circular outcome and thereby illustrate a
wider range of cylindrical models available in the literature. All four
cylindrical models contain a set of $q$ predictors $\boldsymbol{x} = x_1, \dots,
x_q$ and $p$ predictors $\boldsymbol{z} = z_1, \dots, z_q$ for the linear and
circular outcomes, \(Y\) and \(\Theta\), respectively. The first two models are
based on a construction by @mastrantonio2015bayesian, while the other models are
extensions of the models from @abe2017tractable and @mastrantonio2018joint. The
four cylindrical  models are introduced separately in the subsections below.
However, to provide a more succinct overview and comparison of the four models,
\ref{TableModels} gives an overview of the similarities and differences between
the models.

@@

In the Supplementary Material we have described the starting values for the MCMC
procedures for the CL-PN, CL-GPN and GPN-SSN models, hence it remains to specify
the starting  values for the maximum likelihood based Abe-Ley model: \(\eta_0 =
0.9, \eta_1 = 0.9, \nu_0 = 0.9, \nu_1 = 0.9, \kappa = 0.9, \alpha = 0.9, \lambda
= 0\). The initial number of iterations for the three MCMC samplers was set to
2000. After convergence checks via traceplots we concluded that some of the
parameters of the GPN-SSN model did not converge. Therefore we set the number of
iterations of the MCMC models to 20,000 and subtracted a burn-in of 5000 to
reach convergence (the geweke diagnostics show absolute z-scores over 1.96 in
6\% of the estimated parameters). Note that we choose the same number of
iterations for all three models models estimated using MCMC prodedures to make
their comparison via the PLSL as fair as possible. Lastly, the predictor
\verb|SE| was centered before inclusion in the analysis as this allows the
intercepts to bear the classical meaning of average behavior.\newline
\indent Tables \ref{tab:estCLGPN}, \ref{tab:estAL} and \ref{tab:estCLGPNM} show
the results for the four cylindrical models that were fit to the teacher data.
For the models estimated using MCMC methods, CL-PN, CL-GPN and GPN-SSN we show
descriptives of the posterior of the estimated parameters (posterior mode and
lower and upper bound of the 95\% highest posterior density (HPD) interval). For
the Abe-Ley model we show the maximum likelihood estimates of the parameters. To
compare the results of the four models we focus on the following aspects: the
estimated average scores (intercept) on the type and strength of interpersonal
behavior (1), the effect of self-efficacy on the type and strength of
interpersonal behavior (2), the dependence between the type and strength of
interpersonal behavior (3) and the model fit (4).

@@

further covariates on data from the circumplex. Furthermore, in addition to
being able to assess the influence of covariates, the cylindrical models also
provide information about the dependence between the type and strength of
interpersonal behavior. We found that stronger behavior is associated with those
types of behavior that show higher scores on the Communion dimension (and also
on the Agency dimension). This translates to stronger behavior being shown by
teachers whose type of interpersonal behavior ranges between 0$^\circ$ and
90$^\circ$, the 'helpful' and 'directing' subtypes according to Figure
\ref{QTI}. \textcolor{red}{Is dit nieuw/zijn er papers die hier iets over
zeggen?}\newline
\indent As mentioned in the introducion data from the interpersonal circumplex

>>>>

author: 
  - name          : "Jolien Cremers"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "joliencremers@gmail.com"
  - name          : "Helena J.M. Pennings"
    affiliation   : "2,3"
  - name          : "Christophe Ley"
    affiliation   : "4"

@@

authornote: |

@@

In social sciences the use of cylindrical data is very common. Such data consist
of a linear and a circular component. @gurtman2011reasoning refers to such data
as vectors, with a directional measure (i.e., the circular component) and a
measure indicating the magnitude (i.e., linear component). Many established
models in psychology are often referred to as circular or circumplex models, but
those models are cylindrical. Examples of such cylindrical models are the
interpersonal circle/circumplex
[@leary1957;@wiggins1996history;@wubbels2006interpersonal], the circumplex of
affect [@russell1980circumplex], the circumplex of human emotion
[@plutchik1997general] or the model of human values [@schwartz1992values].
\newline
\indent Also, many of the more recent types of data that are studied in psychology are
cylindrical. For example, research on human navigation uses data where distance
(i.e., linear component) and direction (i.e., the linear component) are of
interest [@chrastil2017rotational] or in eye-tracking, the saccade data also
consist of both the direction (i.e., the circular variable) and the duration
(i.e., the linear variable)(e.g., @rayner200935th). Apart from the social
sciences data with a circular and linear outcome more commonly occur in
meteorology [@garcia2013exploring], ecology [@garcia2014test] or
marine research [@lagona2015hidden]
\newline
\indent Up until now researchers studying cylindrical data had to rely on linear
statistical methods to analyze their research results. However, lately more and more
of these researchers acknowledge that linear methods are not sufficient and call
for new methods [@gurtman2011reasoning; @pennings2017complexity;
@wright2009integrating] that take into account both the circular and the linear
component of these data.
\newline
\indent The aim in the present paper is twofold. Firstly, we we intend to fill
the above mentioned gap in the literature by showing the use of cylindrical
models can benefit the analysis of circumplex data and cylindrical data in
psychology in general. More specifically we will show these benefits for
interpersonal teacher data from the field of educational psychology. Apart from
modelling the dependence between the linear and circular component of a
cylindrical variable we would also like to predict the two components from a set
of covariates in a regression model. Our second aim therefore is to adapt
several existing cylindrical models in such a way that they include a regression
of both the linear and circular component of a cylindrical variable onto a set
of covariates. From now on we will therefore refer to the components of the
cylindrical variable as outcome components. These adapted circular models are
then used  to analyse the teacher data.

@@

Data that consist of a variable and a circular variable is called
cylindrical data. A circular variable is different from the linear variable in
the sense that it is measured on a different scale. Figure \ref{circline} shows
the difference between a circular scale (right) and a linear scale (left). The
most important difference is that on a circular scale the datapoints 0\(^\circ\)
and 360\(^\circ\) are connected and in fact represent the same number while on a
linear scale the two ends, \(-\infty\) and \(\infty\), are not connected and
consequently the values 0\(^\circ\) and 360\(^\circ\) are located on different
places on the scale. Both circular data and cylindrical data require special
analysis methods due to this periodicity in the scale of a circular variable
(see e.g. @fisher1995statistical for an introduction to circular data and
@mardia2000directional, @jammalamadaka2001topics and @ley2017modern for a more
elaborate overview).\newline
\indent In the literature, several methods have been put forward to model the
relation between the linear and circular component of a cylindrical variable.
Some of these are based on regressing the linear component onto the circular
component using the following type of relation: \[y = \beta_0 +
\beta_1*\cos(\theta) + \beta_2*\sin(\theta)+ \epsilon,\] where \(y\) is the
linear component and \(\theta\) the circular component [@mardia1978model;
@johnson1978some; @mastrantonio2015bayesian]. Others model the relation in a
different way, e.g. by specifying a multivariate model for several linear and
circular variables and modelling their covariance matrix
[@mastrantonio2018joint] or by proposing a joint cylindrical distribution. For
example, @abe2017tractable introduce a cylindrical distribution based on a
Weibull distribution for the linear component and a sine-skewed von Mises
distribution for the circular component and link these through their respective
shape and concentration parameters. However, none of the methods that have been
proposed thus far include additional covariates onto which both the circular and
linear component are regressed.

@@

(vertical) and Communion (horizontal), see Figure \ref{QTI}. This space is
called the interpersonal circle/circumplex (IPC) and different parts of this
space are characterized by different teacher behavior, e.g. 'helpful' or
'uncertain'. The IPC is ``a continuous order with no beginning or end''
[@gurtman2009exploring, p. 2]. We call such ordering a circumplex ordering and
the IPC is therefore often called the interpersonal circumplex. The ordering

@@

scores. Agency and Communion are measured on a scale from 1 to 5 and for
analysis purposes they are later converted to scale ranging from -1 to 1.  Their
respective 0 scores form the origin. Scaling is however only considered an issue
in those instances where cylindrical data is derived from measurements in
bivariate space.}, where \(A\) represents a score on the Agency dimension and
\(C\) represents a score on the Communion dimension. Note that when placing a
unit circle on Figure \ref{QTI} we see that the Agency dimension is related to
the sine of the circular score and the Communion dimension is related to the
cosine of the circular score.

@@

the two dimensions Agency and Communion instead of both dimensions separately.
This provides us with new insights compared to a separate analysis of the two
dimensions that is standard in the literature (see e.g.
@pennings2018interpersonal, @wright2009integrating or
@wubbels2006interpersonal). There is however one main drawback: when
two-dimensional data are converted to the circle we lose some information,
namely the length of the two-dimensional vector \((A, C)^t\), \emph{i.e.}, its
Euclidean norm \(\mid\mid (A, C)^t \mid\mid\). This length represents the
strength of the interpersonal behavior a teacher shows towards his/her students.
In a cylindrical model this strength (the linear outcome) can be modeled
together with the type of interpersonal behavior of a teacher (the circular
outcome). This leads to an improved analysis of interpersonal circumplex data,
over either analyzing the two dimensions separately or using a circular model,
because we take all information, circular and linear, into account. In the next
section we introduce several cylindrical models that can be used to analyze the
teacher data. First however we will provide descriptives for the teacher data.

@@

The teacher data was collected between 2010 and 2015 and contains several
repeated measures on the IPC of 161 teachers. Measurements were obtained using
the QTI and taken in different years and classes. For this paper we only
consider one measurement, the first occasion (2010) and largest class if data
for multiple classes were available. This results in a sample of 151 teachers.
In addition to the type of interpersonal behavior (IPC), the circular outcome,
and the strength of interpersonal behavior (IPC strength), the linear outcome, a
teachers' self-efficacy (\verb|SE|) concerning classroom management is used as
covariate in the analysis. In previous research, in psychology and education it
has been shown that higher self-efficacy is related to the quality of
interpersonal interactions [@locke2007selfefficacy; @want2018selfefficacy].
After listwise deletion of missings ($3$ in total, only for the self-efficacy)
we have a sample of 148 teachers. Table \ref{Tableteacherdescriptives} shows
descriptives for the dataset. For the circular variable IPC we show sample
estimates for the circular mean $\bar{\theta}$ and concentration $\hat{\rho}$.
The circular concentration lies between 0, meaning the data is not concentrated
at all \emph{i.e.} spread over the entire circle, and 1, meaning all data is
concentrated at a single point on the circle The population values of these
parameters are usually, and also in this paper, referred to as $\mu$ (circular
location) and $\kappa$ (circular concentration). For the linear variables
(strength IPC and SE) we show sample estimates of the linear mean and standard
deviation (sd). Figure \ref{dataplot} is a scatterplot showing the relation between
the linear and circular outcome of the teacher data for teachers with low SE
(below 1 sd below the mean), average SE (between 1 sd below and 1 sd above the
mean) and high SE (above 1 sd above the mean).

@@

One of the goals of this paper is to show the benefits of cylindrical methods
for the analysis of circumplex data and cylindrical data in psychology in
general. To do so we decide to focus on four cylindrical models. The models were
selected for their relatively low complexity and the ease with which a
regression structure could be incorporated. But also because they show different
ways of modelling the the linear and circular outcome and thereby illustrate a
wider range of cylindrical models available in the literature. The
cylindrical models contain a set of $q$ predictors $\boldsymbol{x} = x_1, \dots,
x_q$ and $p$ predictors $\boldsymbol{z} = z_1, \dots, z_q$ for the linear and
circular outcomes, \(Y\) and \(\Theta\), respectively. The first two models are
based on a construction by @mastrantonio2015bayesian, while the other models are
extensions of the models from @abe2017tractable and @mastrantonio2018joint. The
four cylindrical  models are introduced separately in the subsections below.
However, to provide a more succinct overview and comparison of the four models, Table
\ref{TableModels} gives an overview of the similarities and differences between
the models.

@@

In the Supplementary Material we have described the starting values for the MCMC
procedures for the CL-PN, CL-GPN and GPN-SSN models, hence it remains to specify
the starting  values for the maximum likelihood based Abe-Ley model: \(\eta_0 =
0.9, \eta_1 = 0.9, \nu_0 = 0.9, \nu_1 = 0.9, \kappa = 0.9, \alpha = 0.9, \lambda
= 0\). The initial number of iterations for the three MCMC samplers was set to
2000. After convergence checks via traceplots we concluded that some of the
parameters of the GPN-SSN model did not converge. Therefore we set the number of
iterations of the MCMC models to 20,000 and subtracted a burn-in of 5000 to
reach convergence (the Geweke diagnostics show absolute z-scores over 1.96 in
6\% of the estimated parameters). Note that we choose the same number of
iterations for all three models models estimated using MCMC prodedures to make
their comparison via the PLSL as fair as possible. Lastly, the predictor
\verb|SE| was centered before inclusion in the analysis as this allows the
intercepts to bear the classical meaning of average behavior.\newline
\indent Tables \ref{tab:estCLGPN}, \ref{tab:estAL} and \ref{tab:estCLGPNM} show
the results for the four cylindrical models that were fit to the teacher data.
For the models estimated using MCMC methods, CL-PN, CL-GPN and GPN-SSN we show
descriptives of the posterior of the estimated parameters (posterior mode and
lower and upper bound of the 95\% highest posterior density (HPD) interval). For
the Abe-Ley model we show the maximum likelihood estimates of the parameters. To
compare the results of the four models we focus on the following aspects: the
estimated average scores (intercept) on the type and strength of interpersonal
behavior (1), the effect of self-efficacy on the type and strength of
interpersonal behavior (2), the dependence between the type and strength of
interpersonal behavior (3) and the model fit (4).

@@

further covariates on data from the circumplex. \newline
\indent Furthermore, in addition to being able to assess the influence of
covariates, the cylindrical models also provide information about the dependence
between the type and strength of interpersonal behavior. We found that stronger
behavior is associated with higher scores on the Communion and in some models
also the Agency dimension. This implies that teachers whose type of
interpersonal behavior ranges between 0$^\circ$ and 90$^\circ$, the 'helpful'
and 'directing' subtypes are stronger in their behavior than teachers of the
other subtypes.\newline
\indent As mentioned in the introduction, data from the interpersonal circumplex
